---
title: "Functional_analysis_A2"
output: html_document
---

# Functional Analysis - Aim 2

## MSNDR vs. MSLI

This code performs functional analysis using ggpicrust2 between the groups MSNDR and MSLI.

In order to access the data used in the following code, the folders "qiime2_outputs", and "picrust2_outputs" within the datasets folder (located in the root directory) need to be unzipped.

```{r Loading Libraries}

# install.packages("devtools")
# devtools::install_github("cafferychen777/ggpicrust2")
# install.packages('MicrobiomeStat')
# BiocManager::install("KEGGREST")
# install.packages("GGally")
# install.packages("pheatmap")

library(tidyverse)
library(phyloseq)
library(ggpicrust2)
library(pheatmap)

```

```{r Loading Data}

ps <- readRDS("../../datasets/ps_taxonomy.rds")

metadata_formatted <- ps@sam_data %>% data.frame() %>% rownames_to_column("sample_name")

ko <- read.delim("../../datasets/picrust2_outputs/pred_metagenome_unstrat.tsv",
                 row.names = 1)

```

```{r Wrangling}

### Metadata ###

# Filtering for MS patients with either no dietary restrictions, or with
# ...lactose intolerance only

md_MSNDR_MSLI <- metadata_formatted %>%
  filter(disease == "MS")

# write.csv(md_MSNDR_MSLI, "../../datasets/md_MSNDR_MSLI.csv")

### KO ###

ko_col_names <- colnames(ko)

ko_names_formatted <- ko_col_names %>%
  sapply(str_replace, pattern = "X", replacement = "")

ko_formatted <- `colnames<-`(ko, ko_names_formatted)


ko_MSNDR_MSLI <- ko_formatted %>% select(all_of(md_MSNDR_MSLI$sample_name))

```

```{r Pathway Differential Abundance Analysis (DAA)}

daa_results_df = pathway_daa(abundance = ko_MSNDR_MSLI,
                             metadata = md_MSNDR_MSLI, 
                             group = "LI_status", 
                             daa_method = "LinDA", 
                             select = NULL, reference = NULL)

daa_annotated_results_df = pathway_annotation(pathway = "KO",
                                              daa_results_df = daa_results_df,
                                              ko_to_kegg = TRUE)


# saveRDS(daa_annotated_results_df,'../outputs/daa_annotated_results_df.rds')
daa_annotated_results_df = readRDS('daa_annotated_results_df.rds')

daa_annotated_results_df %>% filter(p_adjust< 0.58) %>% nrow()

```

### Notes:

There were insignificant differences between MSNDR and MSLI in terms of functional pathways. Because of this, they could be not be annotated, and no pathway error bar plot could be generated.

```{r PCA Plot(s)}

pca_MSNDR_MSLI <- pathway_pca(abundance = ko_MSNDR_MSLI,
                            metadata = md_MSNDR_MSLI, 
                            group = "LI_status")

pca_MSNDR_MSLI

# ggsave("../outputs/pca_MSNDR_MSLI.png", plot = pca_MSNDR_MSLI)

```

# ```{r Heatmap}
# 
# sig_features = daa_annotated_results_df %>% 
#   filter(p_adjust < 0.56) %>% 
#   pull('feature') %>% unique()
# 
# ko_relab = ko_MSNDR_MSLI %>% 
#   apply(2,function(x) x/sum(x)) %>% as.data.frame()
# # colSums(ko_relab)
# 
# 
# stats = ko_relab %>% t() %>% 
#   as.data.frame() %>% 
#   select(all_of(sig_features)) %>%
#   cor(method = 'pearson')
# 
# 
# color_palette <- colorRampPalette(c("blue", "white", "red"))(40)
# 
# 
# breaks <- seq(-1, 1, length.out = 41)  
# 
# 
# pheatmap(stats, 
#          clustering_distance_rows = "euclidean",  
#          clustering_distance_cols = "euclidean", 
#          clustering_method = "complete",          
#          color = color_palette, 
#          breaks = breaks,
#          main = '',
#          fontsize_row = 10, 
#          fontsize_col = 10,
#          filename = "KEGG_Heatmap_A2.png",
#          height= 9, width = 9)
# 
# ```

### Results:

-   There are no significant pathways below p-adjusted = 0.5

-   Could not generate an appropriate heatmap

-   No functional differences were observed between HNDR and HLI
